---
title: "Workshop #1: introduction to R and RStudio"
author: "BEPHIG"
date: "`r format(Sys.Date())`"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting started
## Environment
Your current working directory can be found by typing the function `getwd()` into your console.
It is important to know the location of your working directory when importing and exporting data to and from R.
You can set a new working directory with the function `setwd()`.  
But before we do that, now is as good a time as any to talk about functions. 
Both `getwd()` and `setwd()` are functions. 
If you want to learn more about a function (e.g., description; usage; arguments), add a question mark before the function to bring up any relevant documentation in your Help pane.

Give these a try:
```{r, results = 'hide'}
?getwd()
getwd()
```

Based on the information included in the documentation, we can use `setwd()` to change our working directory to another folder.
Before we do that, it might be a good idea to assign (`<-`) the current working directory to a variable in case we need to revert back to it in the future.
```{r}
original_directory <- getwd() 
# alternatively
assign(x = "original_directory", value = getwd())
```

Now, we can change the working directory, allowing us to keep all the files relevant to our analysis in a single folder named 'tmp'.
```{r}
setwd(dir = "/tmp")
```

We can then use `getwd()` to see if the change in working directory has been successful.
```{r, results = 'hide'}
getwd()
```

We can restore the original working directory by using the value stored in the variable `original_directory`.
```{r}
setwd(original_directory)
```

Now try `getwd()` again.
```{r, results = 'hide'}
getwd()
```

Notice that `getwd()` will now return the original working directory. 
It is worth pointing out that the `dir = ` portion can be omitted from `setwd(...)` because the function only accepts one argument.  
We can also omit certain arguments from functions if there are default values. 
For example, bring up the documentation for `rnorm()`. 
```{r}
?rnorm()
```

This function generates `n` random values from a normal distribution. 
If we don't specify the mean and standard deviation (sd) of the normal distribution and provide a single number as an argument, then that number will be treated as the argument for `n` and the mean and sd will default to 0 and 1, respectively. 
```{r}
set.seed(1) # to ensure that we get the same values
rnorm(10)
```

But what if we were interested in generating 10 values from a normal distribution with mean of 0 and sd of 10? 
```{r}
set.seed(1)
rnorm(10, 10)
```

Judging from the output, we can already tell that these ten values were drawn from a normal distribution with a mean of 10. 
In this example, where the arguments are ambiguous, the first `10` is interpreted as the first argument (`n`) and the second `10` is interpreted as the second argument (`mean` instead of `sd`).
Thus, we have to be more specific.
```{r}
set.seed(1)
rnorm(10, sd = 10)
```

Bringing the focus back to the topic of working directories, I would recommend the use of RStudio projects.
I find it helpful in compartmentalize my analyses. 
It also plays nicely with git if you are ever in need of version control. 
Detailed information about Rstudio projects can be found on [this page](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects).