---
title: "Workshop #1: Introduction to R and RStudio"
author: "BEPHIG"
date: "`r format(Sys.Date())`"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting started
First, consider setting the option "Save workspace to .RData on exit" to "Never". 
If you become accustomed to beginning anew each time you launch R or RStudio, then reproducible research will become second nature to you.
To change this setting, navigate to Global Options under Tools. 
The option can be toggled between "Always", "Never", and "Ask" in the General tab.

Once you are ready to proceed, start a new R script with the shortcut Ctrl+Shift+N or Cmd+Shift+N, then continue onto the next section.

## Environment
Your current working directory can be found using the function `getwd()`.
Type this into your script, then run the current line of code using the shortcut Ctrl+Enter or Cmd+Enter.
```{r, results = 'hide'}
getwd()
```

You should see the location of your working directory as the output. 
This location is important when importing and exporting information to and from R.
You can set a new working directory with the function `setwd()`.
But before you do that, now is as good a time as any to talk about functions. 
Both `getwd()` and `setwd()` are functions. 
If you want to learn more about a function (e.g., description; usage; arguments), add a question mark before the function to bring up any relevant documentation in the Help pane.
Type the following as a new line of code in your script, then give it a try:
```{r, results = 'hide'}
?getwd()
```

You should now see an updated pane with documentation about the `getwd()` and `setwd()` functions. 
Based on the information provided in the documentation, you should be able to use `setwd()` to change the working directory to another location.
But before you attempt to change the directory, you are advised to assign (`<-`) the path to the current working directory to an object in the environment.
This will allow you to revert back to the original working directory if you need to do so in the future. 
Again, type the following into a new line, then run the code.
```{r}
original_directory <- getwd() # alternatively: assign(x = "original_directory", value = getwd())
```

> You may have noticed text following `original_directory <- getwd()`.
> Anything follow `#` will be ignored when running that line. 
> Use this to your advantage and document your code so that others can follow along.

Now, you can change the working directory to a folder named 'tmp'.
This allows you to keep all the files relevant to your analysis in a single folder.
```{r}
setwd(dir = "/tmp")
```

You can then use `getwd()` to see if the change in working directory has been successful.
```{r, results = 'hide'}
getwd()
```

You can restore the original working directory by using the value stored in the variable `original_directory`.
```{r}
setwd(original_directory)
```

Give `getwd()` another try.
```{r, results = 'hide'}
getwd()
```

Notice that the output of `getwd()` is now the original working directory. 

It is worth pointing out that the `dir = ` portion can be omitted from `setwd(...)` because the function only accepts one argument.
It is also possible to omit certain arguments from functions if the arguments have default values.
For example, bring up the documentation for `rnorm()`. 
```{r}
?rnorm()
```

This function generates `n` random values from a normal distribution. 
If you choose not to specify the mean and standard deviation (sd) of the normal distribution and provide only a single number as an argument, then that number will be treated as the argument for `n` and the mean and sd will default to values of 0 and 1, respectively. 
```{r}
set.seed(1) # to ensure that the same values are generated despite the supposed randomness of rnorm()
rnorm(10)
```

But what if you were interested in generating ten values from a normal distribution with mean of 0 and sd of 10? 
```{r}
set.seed(1)
rnorm(10, 10)
```

Judging from the output, it is obvious that these ten values were drawn from a normal distribution with a mean of 10. 
In this example where the arguments are ambiguous, the first `10` is used as the first argument (`n`) and the second `10` is used as the second argument (`mean` instead of `sd`).
Thus, you will have to be more specific.
```{r}
set.seed(1)
rnorm(10, sd = 10)
```

Circling back to the topic of working directories, consider using RStudio projects to compartmentalize analyses instead. 
When you click on a .Rproj file within your file explorer, you will have already set your working directory to the root folder.
RStudio projects have the added bonus of playing nicely with git/subversion if you are ever in need of version control. 
Detailed information about RStudio projects can be found on [here](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects).

While the following is unnecessary if you had decided to turn off the saving of the workspace to .Rdata on exit, try making it a habit to rid the environment of objects that serve no purpose within your scripts. 
First, you can find out what is actually in the environment using the function `ls()`.
```{r}
ls()
```

The path to the location of the default working directory (i.e., `original_directory`) is not something that you will need to export from R. 
If you were to save the script that you have been working on thus far, you can easily re-create this object as needed. 
Thus, you can remove it from the environment with `rm()` if you intend to start work on a new script.
```{r}
rm(original_directory)
```

You could have also cleared everything from the environment with the following line of code:
```{r}
rm(list = ls())
```

Ctrl+Shift+F10 or Cmd+Shift+F10 can also be used to restart the R session. 
This shortcut is especially helpful because it will also detach/unload any packages you may have attached/loaded previously.

If you were to send this script to someone else, they should be able to replicate each and every step of this demonstration. 

When you are ready to move on, create a dedicated RStudio project and a new R script, then jump into the basics.

## Basics
You can complete simple tasks without relying on the use of functions.
```{r}
1+2
8-4
2*6
49/7
5^2
```

You can also use `==`,`>`, `>=`, etc. (to bring up documentation, try ?`==`) to evaluate if something is TRUE or FALSE.
```{r}
2+2==4
2 + 2 == 5
21>20
21 > 21
21>= 21
```

Adding `!` in front of anything in R acts as a sort of logical negation.
```{r}
2+2==4
2 + 2 != 4
TRUE
!(TRUE)
```

Logical constants can be evaluated in tandem with a combination of `&` ("and") or `|` ("or"). 
```{r}
?`|` # to bring up the relevant documentation
1+1==2 & 2+2==4
TRUE & TRUE
TRUE & FALSE
!FALSE & !FALSE
T | T
F | T
F | F
T | T & F
(T | T) & F

# note: in these contexts, 0 is FALSE and 1 is TRUE
0 == F & 1 == T
```

One application where you might find this helpful is in situations where you want to do something different based on whether the result of the test.
```{r}
ifelse(test = T, yes = "important", no = "not important")
ifelse(F, "important", "not important")
```

As an aside, if you need to make use of the exclusive OR, check out `xor()`:
```{r}
xor(T,F)
xor(F,T)
xor(F,F)
xor(T,T)
```

If you were interested in what is "under the hood" of `xor()`, you can run `xor` (notice the lack of parentheses) and inspect the output.
```{r}
xor
```

Now, it's time to work with some real data.

### Importing data
You can import all sorts of data into R, but some file types may require packages not included with the base version of R.
One popular package for data manipulation is `dplyr`. 
More information on this package can be found [here](https://dplyr.tidyverse.org/).

#### Installing packages not included in base R
You can install `dplyr` (or any other package in CRAN) with the following line of code: `install.packages("dplyr", dependencies = T)`.
The argument, `dependencies = T`, is recommended as this will install any ancillary packages 

You are encouraged to enter and run this line of code in your console because you will only need to install this package once.
It is also common courtesy not to force the installation of a package onto an unsuspecting person running your code. 
They will know soon enough if they are missing a package because they will eventually run into errors. 

Once installed, use the following line to attach the package.
```{r}
library("dplyr")
```

Please also install a sister-package: `readr` (i.e., `install.packages("readr", dependencies = T)`), then attach the package as before.
```{r}
library("readr")
```

Why should you go through the trouble of installing additional packages?
If you intend to use R for data analysis, then this will likely be the first of many packages that you'll have to install to take advantage of R as a statistical language. 

`dplyr` and `readr` are just two of the many packages included in the `tidyverse` suite.
You are encouraged to install `tidyverse` (in the same manner described for `dplyr` and `readr`; i.e., as if it were its own package) on your own time.
While these packages are not strictly necessary for data manipulation, there are many useful functions designed to make your life much simpler. 
Over time, you may discover other packages that work better for you. 

#### Importing data (continued...)
Without further ado, download the following csv file containing data about measles vaccination rates from [rfordatascience's tidytuesday GitHub project](https://github.com/rfordatascience/tidytuesday). 
The original source of these data is [The Wall Street Journal](https://github.com/WSJ/measles-data).
```{r}
measles <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-25/measles.csv')
```

Notice the assignment of the downloaded data to an object termed `measles`. 
This is important because you will be working with the data. 

What exactly are you looking at?
```{r}
str(measles)
```

Based on this output, you can state that you are looking at a tibble with 16 columns and 66,113 rows. 
You can also say that you're looking at a tibble, but what is a tibble? 
A tibble is just a fancier data frame (see [here](https://r4ds.had.co.nz/tibbles.html#tibbles-vs.-data.frame) for a more detailed discussion).

You can summarize the data by using the `summary()` function:
```{r}
summary(measles)
```

Don't worry too much about the output for now. 
Let's take a closer look at the tibble:
```{r}
View(measles)
```

View (with a capital V) opens the tibble (or data frame) in a separate window, allowing you to inspect the different rows and columns.
At a glance, this tibble or "fancy" data frame, looks, more or less, like a well-maintained spreadsheet with each column representing a different variable and each row representing a different observation. 
This is, ideally, how you would structure your data when performing analyses.

Taking a step back, you had only previously worked with single values (recall: `1+2`; `8-4`; etc.). 
How did you get from single values to an entire tibble's worth of data?

The answer has to do with different data structures. 
One of the simplest structures is a vector. 
A vector has to contain information of the same type. 
If you were to inspect the output of `1+2` you would see the following:
```{r}
1+2
str(1+2)
```

This is simply a vector containing 1 numeric value. 

You can have longer numeric or integer vectors. 
For instance:
```{r}
1:5 # the colon generates regular sequence from 1 to 5 (see ?`:` for more information) 
str(1:5)
```

This is a vector containing 5 integers. 
You could re-create the same vector with the following line of code:
```{r}
int_v <- c(1, 2, 3, 4, 5) # note the use of c() as a means of forming a vector
int_v
str(int_v)
```

You can have vectors containing only logical values.
For instance:
```{r}
set.seed(1)
logic_v <- sample(c(T, F), 10, T) # refer to ?sample if you would like to make sense of these arguments
logic_v
str(logic_v)
```

You can have vectors containing ordinal data.
For instance:
```{r}
factor_v <- factor(x = c("unlikely", "very likely", "impossible", "certainty", "unlikely"), # see >factor for more information
                   levels = c("impossible", "unlikely", "very likely", "certainty"))
factor_v
str(factor_v)
```

You can also have vectors containing character data with no implicit ordering.
For instance:
```{r}
(character_v <- c("yellow", "orange", "green", "purple")) # by adding parentheses around the assignment, you can also get an output at the same time
str(character_v)
```

What's most important to grasp is that a vector with different types of data will invariably be made into a character vector.
For instance:
```{r}
(mixed_v <- c(1, TRUE, "character", 5.43, pi))
str(mixed_v)
```

Also important is that you recognize what is going on here with two vectors of different lengths:
```{r}
v1 <- c(2, 7, 1, 0)
v2 <- c(4, 1)
v1 + v2
```

Notice that when adding the two vectors, the first elements of each vector are summed together, the second elements of each vector are summed together, the third element of `v1` and the first element of `v2` is summed together, and the fourth element of `v1` and the second element of `v2` is summed together. 
This is a consequence of the vectorization of many functions in R. 
This can be used to your advantage, but can also have unintended outcomes, so watch out! 
You will also run into problems if your two vectors are of unequal length and the length of the longer vector is not a multiple of the length of the shorter vector.

Putting everything together, you can conceptualize a data frame (or a tibble) as a collection of vectors. 
Each of these vector occupies a column and the vectors are of equal length. 

You can manually put together a small data frame with vectors using the `data.frame()` function. 
Data were randomly sampled in this fictitious example:
```{r}
sample_df <- data.frame("participant" = c(1:10),
                        "age" = sample(18:65, 10, T),
                        "gender" = sample(c("M", "F"), 10, T),
                        "outcome" = sample(c(0, 1), 10, T, prob = c(0.3, 0.7)),
                        "location" = "Ottawa, CA")  # notice how this is only a vector of length 1, but that's OK because of vectorization!
sample_df
str(sample_df)
```

And to prove that a tibble is just a fancy data frame:
```{r}
sample_tibble <- as_tibble(sample_df)
sample_tibble
str(sample_tibble)
```

### Wrapping up
This more or less wraps up the first workshop. 
You should export the measles data as the second workshop will delve deeper into data indexing and manipulation. 
Use the following to keep a local copy of the data around: 
```{r}
write_csv(measles, path = "measles.csv") # the specified path will create a data folder within your current working directory  
```

To be assured that this was exported correctly, import the csv file, this time from your drive, then compare it to the version downloaded from the internet.
```{r}
measles_local <- read_csv("measles.csv")
identical(measles_local, measles)
```

Everything checks out! 
Hope you learned something new.

Consider installing `tidyverse` before the next workshop.
See you next time! 